---
title: "Seasonal Rhythms: Exploratory Analysis of Human Behavior"
author: Poorva Patel
Date: April 22,2024
output: html_notebook
---

Abstract: Globally, criminal activity varies in variation, volume, and frequency throughout the year. The rhythm of the seasons appears to play a role in shaping human behavior, as evidenced by varying patterns of crime rates across different times of the year, also known as crime seasonality. Crime seasonality has been studied since the mid-nineteenth century as it is an important source for predictability of crimes. This exploratory data analysis (EDA) investigates the relationship between the four seasons and crime incidences, seeking to uncover patterns that could offer insights of human behavior. This project will use a combined crime dataset from New Jersey and Massachusetts from 2018-2021 with a focus on violent crime, property crime, and motor vehicular accidents as identifiers of human behavior. These include murder, rape, aggravated assault, motor vehicle accidents, and burglary. The results of the EDA will allow for understanding regarding the effects of seasonality on human behavior. Further understanding of crime seasonality serves as insight for law enforcement and policy makers to develop targeted strategies aimed at crime prevention and public safety enhancement, with the goal to improve community well-being.  
         Keywords: Exploratory Data Analysis, Crime Seasonality, Human Behavior

```{r}
Load libraries
library(dplyr)
library(tidyverse)
library(readr)
library(ggplot2)
library(lubridate)
library(stringr)
```

```{r}
#Read in cleaned dataset
Behavior <- read.csv("seasonal.csv")
```

```{r}
#Preview dataset
glimpse(Behavior)

#Look at Central Tendencies
summary(Behavior)
```
```{r}
#Create seasons by grouping months
Behaviors <- Behavior %>%
  mutate(Season = case_when(
    Month %in% c("December", "January", "February") ~ "Winter",
    Month %in% c("March", "April", "May") ~ "Spring",
    Month %in% c("June", "July", "August") ~ "Summer",
    Month %in% c("September", "October", "November") ~ "Autumn", 
))
```

```{r}
#Analyze which season has the highest incidents of Rape
rape_season <- Behaviors %>%
  group_by(Season) %>%
  summarise(RapeIncidents = sum(rape_total, na.rm = TRUE)) %>%
  arrange(desc(RapeIncidents))
top_rape_season <- rape_season[1, ] 
print(top_rape_season)

###Bar graph for Rape
ggplot(rape_season, aes(x = Season, y = RapeIncidents, fill = Season == "Summer")) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange3")) +
  labs(
    title = "Seasonal Incidents of Rape",
    x = NULL,
    y = "Number of Rapes"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), # Center the title
    panel.grid.major = element_blank(),     # Remove major grid lines
    panel.grid.minor = element_blank(),     # Remove minor grid lines
    legend.position = "none" ) + #take away key
  geom_text(aes(label = RapeIncidents), vjust = -0.3, color = "white")

```
```{r}
#Analyze which season has the highest incidents of crash incidents regarding Motor Vehicles
just_crash_season <- Behaviors %>%
  group_by(Season) %>%
  summarise(JustCrashIncidents = sum(crash_total, na.rm = TRUE)) %>%
  arrange(desc(JustCrashIncidents))
top_just_crash_season <- just_crash_season[1, ] 

print(top_just_crash_season)



###Bar graph for Crashes
ggplot(just_crash_season, aes(x = Season, y = JustCrashIncidents, fill = Season == "Autumn")) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange3")) +
  labs(
    title = "Seasonal Incidents of Crashes",
    x = NULL,
    y = "Number of Crashes"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), 
    panel.grid.major = element_blank(),    
    panel.grid.minor = element_blank(),     
    legend.position = "none" ) + 
  geom_text(aes(label = JustCrashIncidents), vjust = -0.3, color = "white")
```
```{r}
#Analyze which season has the highest incidents of Aggravated Assault
assault_season <- Behaviors %>%
  group_by(Season) %>%
  summarise(AssaultIncidents = sum(aggravated_assault_total, na.rm = TRUE)) %>%
  arrange(desc(AssaultIncidents))
top_assault_season <- assault_season[1, ]
print(top_assault_season)

###Bar graph for Aggravated Assault
ggplot(assault_season, aes(x = Season, y = AssaultIncidents, fill = Season == "Summer")) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange3")) +
  labs(
    title = "Seasonal Incidents of Aggravated Assaults",
    x = NULL,
    y = "Number of Aggravated Assaults"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_blank(),     
    panel.grid.minor = element_blank(),     
    legend.position = "none" ) + 
  geom_text(aes(label = AssaultIncidents), vjust = -0.3, color = "white")
```
```{r}
#Analyze which season has the highest incidents of Murder
murder_season <- Behaviors %>%
  group_by(Season) %>%
  summarise(MurderIncidents = sum(murder_total, na.rm = TRUE)) %>%
  arrange(desc(MurderIncidents))
top_murder_season <- murder_season[1, ]
print(top_murder_season)


###Bar graph for Murder
ggplot(murder_season, aes(x = Season, y = MurderIncidents, fill = Season == "Summer")) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange3")) +
  labs(
    title = "Seasonal Incidents of Murder",
    x = NULL,
    y = "Number of Murders"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_blank(),     
    panel.grid.minor = element_blank(),     
    legend.position = "none" ) + 
  geom_text(aes(label = MurderIncidents), vjust = -0.3, color = "white")
```
```{r}
#Analyze which season has the highest incidents of Burglary 
burglary_season <- Behaviors %>%
  group_by(Season) %>%
  summarise(BurglaryIncidents = sum(burglary_total, na.rm = TRUE)) %>%
  arrange(desc(BurglaryIncidents))
top_burglary_season <- burglary_season[1, ]
print(top_burglary_season)


###Bar graph for Aggravated Assault
ggplot(burglary_season, aes(x = Season, y = BurglaryIncidents, fill = Season == "Summer")) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange3" )) +
  labs(
    title = "Seasonal Incidents of Burglary",
    x = NULL,
    y = "Number of Aggravated Assaults"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_blank(),     
    panel.grid.minor = element_blank(),     
    legend.position = "none" ) + 
  geom_text(aes(label = BurglaryIncidents), vjust = -0.3, color = "white")

```
```{r}
#Creating dates for the incidents
Behaviors %>%
  select(Month, starts_with("rape"), starts_with("murder"), 
         starts_with("burglary"), starts_with("crash"), starts_with("aggravated")) %>%
  pivot_longer(cols = c(starts_with("rape"), starts_with("crash"), starts_with("murder"), 
                        starts_with("burglary"), starts_with("aggravated")),
               names_to = "Year",
               values_to = "Count") %>%
  mutate(Year = str_extract(Year, "\\d+"), 
         Year = str_replace(Year, "_", ""), 
         Date = mdy(paste(Month, "01", Year, sep = " ")))


#Analyze which season has the highest crime incidents 
Behavior_aggregated <- Behaviors %>%
  group_by(Season) %>%
  summarise(TotalIncidents = sum(crash_total, burglary_total, murder_total, rape_total, aggravated_assault_total, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Season = factor(Season, levels = c("Winter", "Spring", "Summer", "Autumn")))


####Make line graph to show overall trends
ggplot(Behavior_aggregated, aes(x = Season, y = TotalIncidents, group = 1)) +
  geom_line(color = "steelblue") +
  geom_point(size = 3,color = "steelblue") +
  labs(
    title = "Trend of Crimes by Season",
    x = NULL,
    y = "Total Incidents"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), 
    panel.grid.major = element_blank(),     
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1))
```
